---
title: "Decomposing Series"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decomposing Series}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
#| include: false
library(trendseries)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Decomposing Series

The trend extraction methods covered in the other vignettes return a single
smooth component. Often, though, an economic series is more naturally described
as the sum of three parts: a slow-moving **trend**, a repeating **seasonal**
pattern, and an irregular **remainder**. `decompose_series()` splits a series
into these three components and adds them as columns to the original data frame.

```{r libs}
#| eval: false
library(trendseries)
library(dplyr)
library(tidyr)
```

The theme below is used throughout the vignette for consistent styling.

```{r theme}
#| code-fold: true
library(ggplot2)

theme_series <- theme_minimal(paper = "#fefefe") +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "#2c3e50"),
    strip.text = element_text(color = "#fefefe"),
    axis.ticks.x = element_line(color = "gray40", linewidth = 0.5),
    axis.line.x = element_line(color = "gray40", linewidth = 0.5),
    axis.title.x = element_blank(),
    palette.colour.discrete = c(
      "#2c3e50",
      "#e74c3c",
      "#f39c12",
      "#1abc9c",
      "#9b59b6"
    )
  )
```

## Trend extraction vs decomposition

It is worth being clear about the difference between this function and
`augment_trends()`.

- `augment_trends()` returns **only the trend** (`trend_*` columns). The
  seasonal and irregular movements are simply smoothed away.
- `decompose_series()` returns **all three components**
  (`trend_*`, `seasonal_*`, and `remainder_*`), and they add back up exactly to
  the original series.

Because there is a seasonal component, `decompose_series()` requires seasonal
data: monthly (`frequency = 12`) or quarterly (`frequency = 4`). For annual
series there is nothing seasonal to isolate, so use `augment_trends()` instead.

## A first decomposition

Let's start with the `gdp_construction` dataset, a quarterly index of Brazilian
construction activity that ships with `trendseries`.

```{r gdp-plot}
ggplot(gdp_construction, aes(date, index)) +
  geom_line(lwd = 0.7) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title = "Brazilian construction activity",
    y = "Index"
  ) +
  theme_series
```

Passing the data frame to `decompose_series()` adds three new columns. The
frequency is detected automatically from the date column.

```{r gdp-decomp}
gdp_parts <- gdp_construction |>
  decompose_series(value_col = "index")

gdp_parts
```

The three components are named after the method (`stl` by default):
`trend_stl`, `seasonal_stl`, and `remainder_stl`.

Reshaping to long format makes it easy to plot the original series alongside its
three components.

```{r gdp-long}
gdp_long <- gdp_parts |>
  pivot_longer(
    cols = c(index, trend_stl, seasonal_stl, remainder_stl),
    names_to = "component",
    values_to = "value"
  ) |>
  mutate(
    component = factor(
      component,
      levels = c("index", "trend_stl", "seasonal_stl", "remainder_stl"),
      labels = c("Observed", "Trend", "Seasonal", "Remainder")
    )
  )
```

```{r gdp-facet}
#| code-fold: true
#| fig-height: 6
ggplot(gdp_long, aes(date, value)) +
  geom_line(aes(color = component), lwd = 0.7, show.legend = FALSE) +
  facet_wrap(vars(component), ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title = "STL decomposition of construction activity",
    subtitle = "Observed = Trend + Seasonal + Remainder",
    y = NULL
  ) +
  theme_series
```

## STL decomposition

The default method is **STL** (Seasonal-Trend decomposition via Loess),
implemented with `stats::stl()`. The seasonal component is estimated with a loess
smoother, the trend with an adaptive moving average, and the remainder is
whatever is left over. The defaults (`s.window = "periodic"`, `robust = FALSE`)
suit most economic series with a stable seasonal pattern.

Fine control is available through the `params` list. The two most useful options
are an evolving seasonal pattern and robust fitting.

```{r stl-params, eval = FALSE}
# Allow the seasonal pattern to evolve slowly over time, and downweight outliers
gdp_robust <- gdp_construction |>
  decompose_series(
    value_col = "index",
    params = list(s.window = 13, robust = TRUE)
  )
```

A `s.window` of `"periodic"` forces an identical seasonal shape in every year;
supplying a positive odd integer instead lets that shape drift gradually. Setting
`robust = TRUE` reduces the influence of one-off spikes on the trend and seasonal
estimates.

## Regression decomposition

The second method, `method = "regression"`, fits a single OLS model

$$
y_t = f(t) + s(t) + \epsilon_t
$$

where $f(t)$ is a polynomial in time and $s(t)$ is a set of period (month or
quarter) dummy variables. The trend is the intercept plus the polynomial terms,
the seasonal component is the dummy-variable contribution, and the remainder is
the model residual.

Unlike STL, the regression trend is a smooth global polynomial, which can be a
better description of a series with a steady long-run direction. For this example, we use the `oil_derivatives` dataset, which records monthly oil production in Brazil in the 1995-2006 window.

```{r reg-decomp}
suboil <- oil_derivatives |>
  filter(between(date, as.Date("1995-06-01"), as.Date("2006-12-01"))) |>
  mutate(lprod = log(production))

ggplot(suboil, aes(date, lprod)) +
  geom_line(lwd = 0.7) +
  labs(
    title = "Petroleum Derivatives Production",
    y = "Thousand barrels per day (log scale)"
  ) +
  theme_series
```

The `trend` argument selects the polynomial form: `"linear"` (the default),
`"quadratic"` for accelerating or decelerating growth, or `"cubic"`. Orthogonal
polynomials are used by default for numerical stability.

```{r}
suboil_reg <- suboil |>
  decompose_series(
    value_col = "lprod",
    method = "regression",
    trend = "cubic"
  )

ggplot(suboil_reg, aes(date)) +
  geom_line(aes(y = lprod, color = "Original"), lwd = 0.7) +
  geom_line(aes(y = trend_regression, color = "Trend (reg.)"), lwd = 0.7) +
  labs(title = "Cubic trend") +
  theme_series
```

Assuming the series is trend-stationary, the `remainder_regression` component is the stationary component.

```{r}
ggplot(suboil_reg, aes(date)) +
  geom_line(aes(y = remainder_regression), lwd = 0.7) +
  theme_series
```

It is instructive to compare the two decomposition methods directly. Both methods produce fixed seasonal patterns, but the STL method follows short-term fluctuations more closely, producing a "cleaner" remainder term.

```{r compare-trends}
#| code-fold: true
comparison <- suboil |>
  decompose_series(value_col = "lprod", method = "stl") |>
  decompose_series(value_col = "lprod", method = "regression", trend = "cubic")

comparison_long <- comparison |>
  pivot_longer(
    cols = trend_stl:remainder_regression,
    names_pattern = "(.*)_(.*)",
    names_to = c("component", "method"),
    values_to = "trend"
  ) |>
  mutate(
    component = factor(component, levels = c("trend", "seasonal", "remainder"))
  )

ggplot(comparison_long, aes(date, trend)) +
  # Original series
  geom_line(
    data = suboil,
    aes(y = lprod),
    layout = c(1, 2),
    lwd = 0.5,
    alpha = 0.5
  ) +
  # Components and methods
  geom_line(aes(y = trend, color = component), lwd = 0.7) +
  facet_grid(vars(component), vars(method), scales = "free_y") +
  guides(color = "none") +
  theme_series
```

## Grouped decomposition

Like `augment_trends()`, `decompose_series()` accepts a `group_cols` argument to
decompose several series at once. The data must be in tidy long format. Here we
use the `electricity` dataset, which records monthly electricity consumption for
three sectors.

```{r elec-decomp}
electricity_parts <- electricity |>
  mutate(lvalue = log(value)) |>
  decompose_series(value_col = "lvalue", group_cols = "name_series")

glimpse(electricity_parts)
```

Each group is decomposed independently, and the components are stacked back into
a single data frame.

```{r elec-plot}
#| code-fold: true
#| fig-height: 5
ggplot(electricity_parts, aes(date)) +
  geom_line(aes(y = remainder_stl, color = "Trend"), lwd = 0.8) +
  facet_wrap(vars(name_series), ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  labs(
    title = "Electricity consumption by sector",
    subtitle = "STL trend extracted per group",
    y = "Consumption",
    color = NULL
  ) +
  theme_series
```

## Multiplicative seasonality

In the two previous examples we log-transformed the series to account for multiplicative seasonality. This is a common pattern in macroeconomic time series, where the seasonal variation grows with the level of the series. Note that both methods are **additive** by design: when the seasonal amplitude
grows with the level (a multiplicative pattern), decompose the series on the log
scale instead.

On the log scale the additive identity still holds; exponentiating the
components recovers a multiplicative decomposition where
`trend * seasonal * remainder = value`.

## Summary

- `decompose_series()` splits a seasonal (monthly or quarterly) series into
  trend, seasonal, and remainder components that sum to the original series.
- `method = "stl"` (default) adapts locally and handles evolving seasonality and
  outliers; `method = "regression"` fits a smooth global polynomial trend with
  seasonal dummies.
- Use `group_cols` to decompose many series at once, and log-transform first for
  multiplicative seasonal patterns.
- For a trend-only estimate, or for annual data, use `augment_trends()` instead.
