---
title: "Trend Extraction Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trend Extraction Methods}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(trendseries)
```

This vignette is a catalogue of every trend-extraction method in `trendseries`:
which family it belongs to, when to reach for it, and which parameters it
accepts. For worked examples of specific families, see the companion
[Moving Averages](moving-averages.html) and
[Econometric Filters](econometric-filters.html) vignettes. To split a series
into trend, seasonal, and remainder components instead of extracting a single
smooth trend, see [Decomposing Series](decompose-series.html).

# The two interfaces

Every method is reachable through two functions that share the same engine and
the same parameters:

- `augment_trends()` — pipe-friendly. Takes a `data.frame`/`tibble`, adds
  `trend_{method}` columns, and supports grouped series via `group_cols`.
- `extract_trends()` — takes a `ts`/`xts`/`zoo` object and returns `ts` objects
  (a single `ts` for one method, a named list for several).

```{r}
# Data-frame interface: adds a trend_stl column
head(augment_trends(ibcbr, value_col = "index", methods = "stl"))
```

```{r}
# Time-series interface: returns a ts object
hp_trend <- extract_trends(AirPassengers, methods = "hp")
class(hp_trend)
```

Pass several methods at once to compare them:

```{r}
trends <- augment_trends(
  ibcbr,
  value_col = "index",
  methods = c("hp", "stl", "henderson")
)
head(trends)
```

# The unified parameter system

Rather than exposing every method's idiosyncratic arguments, `trendseries`
routes a small set of *generic* parameters to whichever method-specific option
they correspond to. Sensible, frequency-aware defaults mean you rarely need to
set them.

| Parameter   | Applies to | Meaning |
|-------------|------------|---------|
| `window`    | moving-average methods (`ma`, `wma`, `triangular`, `ewma`, `median`, `gaussian`, `stl`) | Number of observations in the smoothing window. |
| `smoothing` | `hp`, `loess`, `spline`, `ewma`, `kernel`, `kalman` | Amount of smoothing (interpretation varies by method). |
| `band`      | bandpass filters (`bk`, `cf`) | `c(low, high)` cycle periods to keep. |
| `align`     | `ma`, `wma`, `triangular`, `gaussian` | `"center"` (default), `"right"` (causal), or `"left"`. |
| `params`    | all | A named list for any remaining method-specific options. |

```{r}
# A wider HP smoothing and a 12-month moving average, in one call
augment_trends(
  ibcbr,
  value_col = "index",
  methods = c("hp", "ma"),
  smoothing = 1600,
  window = 12
) |>
  head()
```

Frequency-aware defaults adapt to the data: for monthly series the HP smoothing
parameter defaults to `lambda = 14400` and the moving-average window to 12; for
quarterly series, `lambda = 1600` and a window of 4 (Ravn & Uhlig, 2002).

# Method catalogue

`trendseries` ships 20 trend methods across four families.

| Method       | Category       | Description                            | Typical use |
|--------------|----------------|----------------------------------------|-------------|
| `hp`         | econometric    | Hodrick-Prescott filter                | General-purpose business-cycle trend |
| `hamilton`   | econometric    | Hamilton regression filter             | HP alternative without spurious cycles |
| `bn`         | econometric    | Beveridge-Nelson decomposition         | Permanent/transitory split |
| `ucm`        | econometric    | Unobserved components model            | Model-based, stochastic trend |
| `bk`         | bandpass       | Baxter-King bandpass filter            | Isolating a cycle frequency band |
| `cf`         | bandpass       | Christiano-Fitzgerald bandpass filter  | Asymmetric bandpass, uses endpoints |
| `ma`         | moving average | Simple moving average                  | Quick, intuitive smoothing |
| `wma`        | moving average | Weighted moving average                | Smoothing with custom weights |
| `ewma`       | moving average | Exponentially weighted moving average  | Recent-weighted, real-time smoothing |
| `triangular` | moving average | Triangular moving average              | Smoother than a simple MA |
| `median`     | moving average | Median filter                          | Robust to outliers/spikes |
| `gaussian`   | moving average | Gaussian-weighted moving average       | Smooth, bell-weighted average |
| `spencer`    | moving average | Spencer's 15-term moving average       | Classic actuarial graduation |
| `henderson`  | moving average | Henderson moving average               | Trend term inside X-11 seasonal adj. |
| `stl`        | smoothing      | Seasonal-trend decomposition via Loess | Trend from strongly seasonal data |
| `loess`      | smoothing      | Local polynomial regression            | Flexible non-parametric trend |
| `spline`     | smoothing      | Smoothing splines                      | Smooth curve with automatic penalty |
| `poly`       | smoothing      | Polynomial trends                      | Simple global trend shape |
| `kernel`     | smoothing      | Kernel smoother                        | Non-parametric, bandwidth-controlled |
| `kalman`     | smoothing      | Kalman filter/smoother                 | Adaptive trend for noisy series |

## Moving averages

Moving-average methods replace each point with a (possibly weighted) average of
its neighbours. They are fast, transparent, and a good default for exploratory
work. Control the smoothing through `window` (and `align` for causal vs.
centred variants). `ma`, `median`, and `henderson` also accept a *vector* of
windows, returning one trend per window.

```{r}
augment_trends(
  ibcbr,
  value_col = "index",
  methods = "henderson",
  window = c(13, 23)
) |>
  head()
```

See [Moving Averages](moving-averages.html) for the full treatment.

## Smoothing methods

Smoothing methods fit a flexible curve to the data. `stl` and `loess` are
locally adaptive; `spline` and `kernel` trade off fit against smoothness through
a penalty/bandwidth; `poly` imposes a single global shape. The `smoothing`
parameter tunes how aggressively they smooth.

```{r}
loess_trend <- extract_trends(AirPassengers, methods = "loess", smoothing = 0.3)
plot(AirPassengers, col = "grey60", ylab = "Air passengers")
lines(loess_trend, col = "#C53030", lwd = 2)
```

## Econometric filters

These are the workhorses of applied macroeconomics. The Hodrick-Prescott filter
(`hp`) is the most widely used; `hamilton` is a regression-based alternative
that avoids HP's well-known spurious-cycle artefacts; `bn` and `ucm` are
model-based decompositions into permanent and transitory parts.

```{r}
augment_trends(
  gdp_construction,
  value_col = "index",
  methods = c("hp", "hamilton")
) |>
  head()
```

The HP filter has a one-sided (real-time) variant for nowcasting, where future
observations must not influence the current estimate:

```{r}
extract_trends(
  AirPassengers,
  methods = "hp",
  params = list(hp_onesided = TRUE)
) |>
  head()
```

See [Econometric Filters](econometric-filters.html) for details.

## Bandpass filters

Bandpass filters (`bk`, `cf`) keep only the fluctuations whose periodicity falls
inside a chosen band, removing both the long-run trend and high-frequency noise.
Specify the band with `band = c(low, high)` in periods (quarters for quarterly
data):

```{r}
extract_trends(
  AirPassengers,
  methods = "cf",
  band = c(18, 96)
) |>
  head()
```

# Decomposition vs. trend extraction

The methods above estimate a single smooth *trend*. When you instead want to
split a series into **trend + seasonal + remainder**, use `decompose_series()`,
which offers five methods of its own:

| Method        | Engine | Notes |
|---------------|--------|-------|
| `stl`         | `stats::stl()` | Loess-based, robust option available. |
| `regression`  | OLS | Polynomial trend + seasonal dummies. |
| `classic`     | `stats::decompose()` | Classical moving-average; additive or multiplicative. |
| `bsm`         | `stats::StructTS()` | State-space model; components for every point. |
| `seats`       | X-13ARIMA-SEATS | Requires the optional **`seasonal`** package. |

```{r}
decompose_series(gdp_construction, value_col = "index", method = "stl") |>
  head()
```

The dedicated [Decomposing Series](decompose-series.html) vignette covers these
in depth.

# References

Ravn, M. O., & Uhlig, H. (2002). On adjusting the Hodrick-Prescott filter for
the frequency of observations. *The Review of Economics and Statistics*, 84(2),
371–376.
