---
title: "Trend Extraction Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trend Extraction Methods}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

```{r setup}
library(trendseries)
```

This vignette catalogues the trend-extraction methods in `trendseries`: which
family each belongs to, when to reach for it, and which parameters it accepts.
For worked examples of specific families, see the companion
[Moving Averages](moving-averages.html) and
[Econometric Filters](econometric-filters.html) vignettes. To split a series
into trend, seasonal, and remainder components instead of extracting a single
smooth trend, see [Decomposing Series](decompose-series.html).

# The two interfaces

Two functions share the same engine and the same parameters, and either one
reaches every method.

- `augment_trends()` — pipe-friendly. Takes a `data.frame`/`tibble`, adds
  `trend_{method}` columns, and supports grouped series via `group_cols`.
- `extract_trends()` — takes a `ts`/`xts`/`zoo` object and returns `ts` objects
  (a single `ts` for one method, a named list for several).

```{r}
# Data-frame interface: adds a trend_stl column
head(augment_trends(ibcbr, value_col = "index", methods = "stl"))
```

```{r}
# Time-series interface: returns a ts object
hp_trend <- extract_trends(AirPassengers, methods = "hp")
class(hp_trend)
```

Pass several methods at once to compare them.

```{r}
trends <- augment_trends(
  ibcbr,
  value_col = "index",
  methods = c("hp", "stl", "henderson")
)
head(trends)
```

# The unified parameter system

Rather than exposing every method's idiosyncratic arguments, `trendseries`
routes a small set of *generic* parameters to whichever method-specific option
they correspond to. All of them have defaults, so you can leave them unset.

- `window` — number of observations in the smoothing window.
- `smoothing` — amount of smoothing. The scale depends on the method; see
  [Reading `smoothing`](#reading-smoothing).
- `band` — `c(low, high)`, the cycle periods removed by a bandpass filter.
- `align` — `"center"` (default) or `"right"`, which uses only past and
  current observations. `ma` and `wma` also accept `"left"`.
- `params` — a named list for any remaining method-specific option.

```{r}
# A wider HP smoothing and a 5-month moving average, in one call
augment_trends(
  ibcbr,
  value_col = "index",
  methods = c("hp", "ma"),
  smoothing = 1600,
  window = 5
)
```

Some defaults follow the frequency of the series. For monthly data the HP
`lambda` defaults to 129600 and the `ma`, `wma`, and `triangular` windows to
12; for quarterly data, to 1600 and 4. The HP default scales 1600 by the
fourth power of the frequency ratio (Ravn and Uhlig 2002), which gives 6.25
for annual data. This keeps the trend-cycle cutoff near ten years at every
frequency. Many other tools default to 14400 for monthly data; pass
`smoothing = 14400` to match them. The `band` default covers cycles of 1.5
to 8 years: `c(6, 32)` for quarterly data and `c(18, 96)` for monthly.

## Reading `smoothing`

The same number means different things to different methods.

- `hp`: a value above 1 is the `lambda` itself. A value of 1 or less is a
  fraction of the frequency's standard `lambda`, so `smoothing = 0.5` on
  monthly data gives `lambda = 64800`.
- `loess`: the span, the share of observations in each local fit.
- `spline`: the `spar` penalty. Left unset, `spline` picks it by generalised
  cross-validation.
- `ewma`: the weight `alpha` on the newest observation. Pass either `window`
  or `smoothing` to `ewma`, not both.
- `kernel`: a multiplier on the automatically chosen bandwidth.
- `kalman`: the ratio of measurement noise to process noise in a local-level
  model. Larger values give a smoother trend.

# Method catalogue

```{r catalogue, echo = FALSE, eval = TRUE}
info <- trendseries:::.method_info()
params <- trendseries:::.method_params()
catalogue <- merge(info, params, by = "method", sort = FALSE)

mark <- function(x) ifelse(x, "✓", "")

catalogue$window <- ifelse(
  catalogue$window_vector,
  "✓ vector",
  mark(catalogue$window)
)
catalogue$smoothing <- mark(catalogue$smoothing)
catalogue$band <- mark(catalogue$band)
catalogue$align <- mark(catalogue$align)
catalogue$one_sided <- c(always = "✓", option = "optional", no = "")[
  catalogue$one_sided
]
catalogue$method <- paste0("`", catalogue$method, "`")
```

```{r catalogue-table, echo = FALSE, eval = TRUE}
families <- c(
  econometric = "Econometric filters",
  bandpass = "Bandpass filters",
  moving_average = "Moving averages",
  smoothing = "Smoothing methods"
)
columns <- c(
  "method",
  "description",
  "typical_use",
  "window",
  "smoothing",
  "band",
  "align",
  "one_sided"
)

family_blocks <- lapply(names(families), function(family) {
  rows <- catalogue[catalogue$category == family, columns]
  header <- rows[1, ]
  header[] <- ""
  header$method <- paste0("**", families[[family]], "**")
  return(rbind(header, rows))
})
catalogue_table <- do.call(rbind, family_blocks)

knitr::kable(
  catalogue_table,
  row.names = FALSE,
  col.names = c(
    "Method",
    "Description",
    "Typical use",
    "`window`",
    "`smoothing`",
    "`band`",
    "`align`",
    "One-sided"
  ),
  align = c("l", "l", "l", "c", "c", "c", "c", "c")
)
```

Every method also accepts method-specific options through `params`. A few
entries in the table need a note.

- `stl`: `window` sets the **seasonal** window (`s.window`), not the trend
  window. Set the trend window with `params = list(t.window = ...)`.
- `vector`: pass several windows, such as `window = c(13, 23)`, to get one
  trend per window.
- `triangular` and `gaussian` accept `align = "center"` or `"right"` only.
- One-sided methods use only past and current observations, which matters for
  real-time work. `ewma` is always one-sided. The methods marked *optional* are
  one-sided with `align = "right"`, or, for `hp`, with
  `params = list(hp_onesided = TRUE)`.

## Moving averages

Moving-average methods replace each point with a (possibly weighted) average of its neighbours, which makes them a good starting point for exploratory work. Control the smoothing through `window`, and the alignment through `align`. `ma`, `median`, and `henderson` also accept a *vector* of windows, returning one trend per window.

```{r}
augment_trends(
  ibcbr,
  value_col = "index",
  methods = "henderson",
  window = c(13, 23)
)
```

See [Moving Averages](moving-averages.html) for the full treatment.

## Smoothing methods

Smoothing methods fit a flexible curve to the data. `stl` and `loess` are
locally adaptive; `spline` and `kernel` trade off fit against smoothness
through a penalty or a bandwidth; `poly` imposes a single global shape. The
`smoothing` parameter tunes how aggressively they smooth.

```{r}
loess_trend <- extract_trends(AirPassengers, methods = "loess", smoothing = 0.3)
plot(AirPassengers, col = "grey60", ylab = "Air passengers")
lines(loess_trend, col = "#D3742A", lwd = 2)
```

## Econometric filters

These are the typical filters of applied macroeconomics. The Hodrick-Prescott filter
(`hp`) is the most widely used; `hamilton` is a regression-based alternative
designed to avoid the spurious cycles HP can introduce; `bn` and `ucm` are
model-based decompositions into permanent and transitory parts.

```{r}
augment_trends(
  gdp_construction,
  value_col = "index",
  methods = c("hp", "hamilton")
) |>
  head()
```

The HP filter has a one-sided (real-time) variant for nowcasting, where future
observations must not influence the current estimate.

```{r}
extract_trends(
  AirPassengers,
  methods = "hp",
  params = list(hp_onesided = TRUE)
) |>
  head()
```

See [Econometric Filters](econometric-filters.html) for details.

## Bandpass filters

Bandpass filters (`bk`, `cf`) isolate the fluctuations whose period falls
inside a chosen band. The trend they return is the series *minus* that band:
the long-run movement plus any high-frequency noise. To get the band itself,
use `detrend_series()`. Specify the band with `band = c(low, high)`, in
periods of the series (months for monthly data). `bk` leaves missing values
at both ends of the series; `cf` does not.

```{r}
extract_trends(
  AirPassengers,
  methods = "cf",
  band = c(18, 96)
) |>
  head()
```

# Decomposition vs. trend extraction

The methods above estimate a single smooth *trend*. When you instead want to
split a series into **trend + seasonal + remainder**, use `decompose_series()`,
which has methods of its own.

| Method        | Engine | Notes |
|---------------|--------|-------|
| `stl`         | `stats::stl()` | Loess-based, robust option available. |
| `regression`  | OLS | Polynomial trend + seasonal dummies. |
| `classic`     | `stats::decompose()` | Classical moving-average; additive or multiplicative. |
| `bsm`         | `stats::StructTS()` | State-space model; components for every point. |
| `seats`       | X-13ARIMA-SEATS | Requires the optional **`seasonal`** package. |

```{r}
decompose_series(gdp_construction, value_col = "index", methods = "stl") |>
  head()
```

The dedicated [Decomposing Series](decompose-series.html) vignette covers these
in depth.

# References

Ravn, M. O., & Uhlig, H. (2002). On adjusting the Hodrick-Prescott filter for
the frequency of observations. *The Review of Economics and Statistics*, 84(2),
371–376.
