---
title: "Moving Averages for Trend Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moving Averages for Trend Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
#| include: false
library(trendseries)
library(patchwork)
library(dplyr)
library(ggplot2)
library(tidyr)
```

# Moving Averages

Moving averages are one of the most intuitive and widely-used tools for
extracting trends from time series data. The basic idea is simple: average
nearby observations to smooth out random fluctuations.

## Simple example

```{r}
library(trendseries)
library(dplyr)
library(ggplot2)
```

To recreate the plots from this tutorial use `theme_series` below.

```{r}
#| code-fold: true
theme_series <- theme_minimal(paper = "#fefefe") +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "#2c3e50"),
    strip.text = element_text(color = "#fefefe"),
    axis.ticks.x = element_line(color = "gray40", linewidth = 0.5),
    axis.line.x = element_line(color = "gray40", linewidth = 0.5),
    # Use colors
    palette.colour.discrete = c(
      "#2c3e50",
      "#e74c3c",
      "#f39c12",
      "#1abc9c",
      "#9b59b6"
    )
  )
```

Let's start with vehicle production data. This series is packaged with
`trendseries` and shows the amount of vehicles produced in Brazil per month.

```{r vehicles-plot}
# Using the 'vehicles' dataset (ships with trendseries)
vehicles_recent <- vehicles |>
  # Only use data after 2018
  filter(date >= as.Date("2018-01-01"))

ggplot(vehicles_recent, aes(date, production)) +
  geom_line(lwd = 0.7) +
  theme_series
```

Applying `augment_trends()` to the `vehicles` dataset creates a new column
called `trend_ma`.

```{r ma-basic}
# Apply a moving average trend
vehicles_trend <- augment_trends(
  vehicles_recent,
  value_col = "production",
  methods = "ma",
  window = 12
)


vehicles_trend
```

We can visualize this trend using `ggplot2`.

```{r}
ggplot(vehicles_trend, aes(date)) +
  geom_line(aes(y = production, color = "Original"), lwd = 0.6, alpha = 0.8) +
  geom_line(aes(y = trend_ma, color = "Trend: 12-month MA"), lwd = 0.8) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x = NULL, y = NULL, title = "Vehicle Production: Simple Moving Average") +
  theme_series
```

The `augment_trends` function accepts a vector of values for `window`. Each
window size produces its own column: `trend_ma_3`, `trend_ma_6`, `trend_ma_12`,
`trend_ma_24`.

```{r window-comparison}
# Apply different window sizes
windows_to_test <- c(3, 6, 12, 24)

vehicles_trend <- vehicles_recent |>
  augment_trends(
    value_col = "production",
    methods = "ma",
    window = windows_to_test
  )

vehicles_trend
```

For plots with more series, reshaping the data to a "tidy" long format is more
convenient.

```{r}
# Prepare for plotting
plot_data <- vehicles_trend |>
  pivot_longer(
    cols = c(production, starts_with("trend_ma")),
    names_to = "method",
    values_to = "value"
  ) |>
  mutate(
    method = factor(
      method,
      levels = c("production", paste0("trend_ma_", c(3, 6, 12, 24))),
      labels = c(
        "Original",
        "3-month MA",
        "6-month MA",
        "12-month MA",
        "24-month MA"
      )
    )
  )


# Plot
ggplot(plot_data, aes(date, value, color = method)) +
  geom_line() +
  labs(
    title = "Effect of Window Size on Moving Average",
    subtitle = "Larger windows = smoother trends, but slower to react",
    x = NULL,
    y = NULL,
    color = NULL
  ) +
  theme_series
```

Notice how the 24-month MA is very smooth but lags behind changes, while the
3-month MA tracks the data closely but still shows some fluctuation.

`augment_trends` supports different alignments via the `align` parameter.

```{r}
vehicles_trend <- augment_trends(
  vehicles_recent,
  value_col = "production",
  methods = "ma",
  window = 12,
  align = "right"
)

ggplot(vehicles_trend, aes(date)) +
  geom_line(aes(y = production, color = "Original"), lwd = 0.6, alpha = 0.8) +
  geom_line(aes(y = trend_ma, color = "Trend: 12-month MA"), lwd = 0.8) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x = NULL, y = NULL, title = "Vehicle Production: Simple Moving Average") +
  theme_series
```

## Grouped series

Working with multiple time series is straightforward. The `augment_trends`
function accepts a `group_cols` argument to apply methods to each group
independently. The data must be in "tidy" long format. Here we use the
`transit_london_monthly` dataset, which aggregates ridership by Bus and Train
(tube).

```{r}
transit <- transit_london_monthly

ggplot(transit, aes(date_month, journey_monthly, color = transit_mode)) +
  geom_line(lwd = 0.7) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_comma(scale = 1e-6)) +
  labs(
    x = NULL,
    y = "Journeys (million)",
    title = "Transit ridership in London",
    subtitle = "Monthly journey counts averaged across London's transit systems",
    color = NULL
  ) +
  theme_series
```

```{r}
transit_trends <- augment_trends(
  transit,
  date_col = "date_month",
  value_col = "journey_monthly",
  group_cols = "transit_mode",
  methods = "ma",
  window = 12
)

ggplot(transit_trends, aes(date_month, color = transit_mode)) +
  geom_line(aes(y = journey_monthly), lwd = 0.7, alpha = 0.8) +
  geom_line(aes(y = trend_ma), lwd = 0.8) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_comma(scale = 1e-6)) +
  labs(
    x = NULL,
    y = "Journeys (million)",
    title = "Grouped series trends",
    subtitle = "Monthly journey counts averaged across London's transit systems",
    color = NULL
  ) +
  theme_series
```

## Related methods

Other window-based smoothing methods are available in `trendseries`, selected
via the `methods` parameter:

1. Moving median `methods = "median"`.
2. Weighted moving average `methods = "wma"`.
3. Exponentially weighted moving average `methods = "ewma"`.
4. Spencer moving average `methods = "spencer"`.
5. Triangular moving average `methods = "triangular"`.

These different methods can be combined in a single call to `augment_trends()`.

```{r}
transit_trends <- augment_trends(
  transit,
  date_col = "date_month",
  value_col = "journey_monthly",
  group_cols = "transit_mode",
  methods = c("ma", "median", "spencer")
)
```

As with multiple windows, `augment_trends` creates a new column for each method.

```{r}
transit_trends
```

Finally, we can visualize these different trends using `ggplot2`.

```{r}
#| code-fold: true
#| fig-width: 9
#| fig-height: 6
transit_trends <- augment_trends(
  transit,
  date_col = "date_month",
  value_col = "journey_monthly",
  group_cols = "transit_mode",
  methods = c("ma", "median", "spencer")
)

transit_trends_long <- transit_trends |>
  pivot_longer(
    cols = c(starts_with("trend_")),
    names_to = "method",
    names_repair = "unique"
  ) |>
  mutate(
    method = factor(
      method,
      levels = c(
        "trend_ma",
        "trend_median",
        "trend_spencer"
      ),
      labels = c(
        "12-month MA",
        "5-month median",
        "15-term Spencer"
      )
    )
  )

ggplot() +
  geom_line(
    data = transit_trends,
    aes(date_month, journey_monthly),
    lwd = 0.5,
    alpha = 0.8
  ) +
  geom_line(
    data = transit_trends_long,
    aes(date_month, value, color = method),
    lwd = 0.8
  ) +
  facet_wrap(vars(transit_mode, method), ncol = 3) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_comma(scale = 1e-6)) +
  labs(
    x = NULL,
    y = "Journeys (million)",
    title = "Grouped series trends",
    subtitle = "Monthly journey counts averaged across London's transit systems",
    color = NULL
  ) +
  theme_series +
  theme(
    axis.text.x = element_text(angle = 90)
  )
```
