---
title: "Moving Averages for Trend Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moving Averages for Trend Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(trendseries)
library(dplyr)
library(ggplot2)
library(tidyr)
```

# Moving Averages

## Introduction

Moving averages are one of the most intuitive and widely-used tools for
extracting trends from time series data. The basic idea is simple: average
nearby observations to smooth out random fluctuations.

## Moving Averages

A moving average, as the name suggests, calculates the average across time
windows. Typically, the moving average serves an estimate of the trend of the
series. Applying a moving average to a time series $y_t$ produces a new series
$z_t$:

$$z_t = \sum_{j = -k}^{k}a_{j}y_{t+j}$$

In the simplest case, $k = 1$ we have:

$$z_{t} = a_1y_{t-1} + a_2 y_{t} + a_3 y_{t+1}$$

Assuming all weights $a_{j}$ are equal and sum to one, we have the simple moving
average below:

$$z_{t} = \frac{1}{3}y_{t-1} +\frac{1}{3} y_{t} + \frac{1}{3} y_{t+1}$$

This examples shows a symetrical moving average of order 3, where all weights
are equal. Each point in the new series $z_t$ is the average between neighboring
values int the original series. Typically, nearby values tend to be similar; in
practice, this makes the moving average filter very smooth.

It's important to note that it's not possible to calculate this filter both at
the beginning and the of the $y_t$ series. This implies that the new series
$z_t$ has fewer observations than the original $y_t$ series. This is a downside
when using moving average filters.

Symetrical moving averages always have an odd number of terms: in the example
above, $k=1$ resulted in a filter with three terms; if we had used $k=2$ we
would have five terms and so on. It's possible to make non-symmetric moving
averages varying the window size. The equation below shows a filter that sums
the last two observations (t-2, t-1) and takes the average with the current
observation (t) and the next observation (t+1).

$$z_{t} = \frac{1}{4} y_{t-2} + \frac{1}{4} y_{t-1} + \frac{1}{4} y_{t} + \frac{1}{4} y_{t+1}$$

To make this moving average symmetric, we can calculate another moving-average:

Não existe forte "contraindicação" sobre o uso de médias móveis não-simétricas.
Vale notar, contudo, que é relativamente fácil transformar uma média móvel
não-simétrica em uma média móvel simétrica. Vamos tirar uma média móvel sobre a
média móvel acima:

$$\begin{align}
x_{t} & = \frac{1}{2}z_{t} + \frac{1}{2}z_{t+1} \\
x_{t} & = \frac{1}{2}(\frac{1}{4}(y_{t-2} + y_{t-1} + y_{t} + y_{t+1}) + \frac{1}{4}(y_{t-1} + y_{t} + y_{t+1} + y_{t+2})) \\
x_{t} & = \frac{1}{8} y_{t-2} + \frac{1}{4} y_{t-1} + \frac{1}{4} y_{t} + \frac{1}{4} y_{t+1} + \frac{1}{8} y_{t+2}
\end{align}$$

The final series is a symmetrical moving average with smaller weights on each
end of the series. This is a moving average of order 2x4.

## Examples

```{r}
library(trendseries)
library(dplyr)
library(ggplot2)

theme_series <- theme_minimal(paper = "#fefefe") +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    # Use colors
    palette.colour.discrete = c(
        "#2c3e50",
        "#e74c3c",
        "#f39c12",
        "#1abc9c",
        "#9b59b6"
    )
  )
```

Let's start with vehicle production data.

```{r ma-basic}
subvehicles <- vehicles |>
  rename(value = production) |>
  filter(date >= as.Date("2018-01-01"))

subvehicles <- augment_trends(
  subvehicles,
  methods = "ma",
  window = 12
  )

# Create vehicles_recent for subsequent examples
vehicles_recent <- vehicles |>
  filter(date >= as.Date("2018-01-01"))
```

```{r}
plot_data <- subvehicles |>
  tidyr::pivot_longer(cols = -date, names_to = "series") |>
  mutate(
    series = factor(
      series,
      levels = c("value", "trend_ma"),
      labels = c("Original", "12-month MA")
    )
  )

ggplot(plot_data, aes(x = date, y = value, color = series)) +
  geom_line(linewidth = 0.8) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Vehicle Production: Simple Moving Average",
    subtitle = "12-month window smooths out month-to-month variation",
    x = NULL,
    y = "Production (thousands of units)",
    color = NULL
  ) +
  theme_series
```

The moving average clearly shows the underlying trend by filtering out the
month-to-month noise.

To compare different window sizes `augment_trends` can accept a vector of values
to `window`. Note that this is currently only available for methods `ma` and
`median`, that is moving averages and moving medians.

```{r window-comparison}
# Apply different window sizes
windows_to_test <- c(3, 6, 12, 24)

vehicles_trend <- vehicles_recent |>
  augment_trends(
    value_col = "production",
    methods = "ma",
    window = windows_to_test
  )

# Prepare for plotting
plot_data <- vehicles_trend |>
  select(date, production, starts_with("ma_")) |>
  pivot_longer(
    cols = c(production, starts_with("ma_")),
    names_to = "method",
    values_to = "value"
  ) |>
  mutate(
    method = factor(
      method,
      levels = c("production", "ma_3m", "ma_6m", "ma_12m", "ma_24m"),
      labels = c(
        "Original",
        "3-month MA",
        "6-month MA",
        "12-month MA",
        "24-month MA"
      )
    )
  )

# Plot
ggplot(plot_data, aes(x = date, y = value, color = method)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Effect of Window Size on Moving Average",
    subtitle = "Larger windows = smoother trends, but slower to react",
    x = "Date",
    y = "Production (thousands of units)",
    color = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Notice how the 24-month MA is very smooth but "lags" behind changes, while the
3-month MA tracks the data closely but still shows some fluctuation.

### Window Size Guidelines

### Understanding Alignment: Center vs Right vs Left

Moving averages can be calculated with different **alignments**, which
determines which observations are used to calculate each point. This is a
critical choice that affects both the trend's properties and when NAs appear in
the result.

1.  **Center alignment** (default): Uses observations both before and after each
    point
    - Most common for general trend extraction
    - Produces NAs at both the beginning and end of the series
    - **Non-causal**: uses future information
2.  **Right alignment** (causal): Uses only past observations
    - Critical for real-time analysis and forecasting
    - Produces NAs only at the beginning
    - **No look-ahead bias**: suitable for backtesting strategies
    - Also called "trailing" or "backward-looking"
3.  **Left alignment** (anti-causal): Uses only future observations
    - Rarely used in practice
    - Produces NAs only at the end
    - Useful for specific smoothing applications

**Use center alignment when:** - Doing historical analysis where all data is
available - You want the smoothest possible trend - The symmetric window makes
sense for your application

**Use right alignment when:** - Building forecasting models (avoid look-ahead
bias) - Backtesting trading strategies or economic indicators - Analyzing data
in real-time (can't use future data) - Need causal filters for time series
econometrics

**Use left alignment when:** - Specific smoothing applications that need
forward-looking averages - Very rarely used in economic analysis

### Visualizing Different Alignments

Let's compare the three alignments using vehicle production data:

```{r align-comparison}
# Apply 12-month moving average with different alignments
vehicles_align <- vehicles_recent |>
  augment_trends(
    value_col = "production",
    methods = "ma",
    window = 12,
    align = "center"
  ) |>
  rename(trend_center = trend_ma)

# Add right alignment
vehicles_align <- vehicles_align |>
  augment_trends(
    value_col = "production",
    methods = "ma",
    window = 12,
    align = "right"
  ) |>
  rename(trend_right = trend_ma)

# Add left alignment
vehicles_align <- vehicles_align |>
  augment_trends(
    value_col = "production",
    methods = "ma",
    window = 12,
    align = "left"
  ) |>
  rename(trend_left = trend_ma)

# Prepare for plotting
plot_data <- vehicles_align |>
  select(date, production, starts_with("trend_")) |>
  pivot_longer(
    cols = starts_with("trend_"),
    names_to = "alignment",
    values_to = "value"
  ) |>
  mutate(
    alignment = case_when(
      alignment == "trend_center" ~ "Center (default)",
      alignment == "trend_right" ~ "Right (causal)",
      alignment == "trend_left" ~ "Left (anti-causal)"
    ),
    alignment = factor(
      alignment,
      levels = c("Center (default)", "Right (causal)", "Left (anti-causal)")
    )
  )

# Plot
ggplot(plot_data, aes(x = date, y = value, color = alignment)) +
  geom_line(linewidth = 0.9, alpha = 0.8) +
  labs(
    title = "Moving Average Alignment Comparison",
    subtitle = "12-month window with different alignments",
    x = "Date",
    y = "Production (thousands of units)",
    color = "Alignment"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Notice how: - **Center** is smoothest and symmetric - **Right** lags behind
center (uses only past data) - **Left** leads ahead of center (uses only future
data)

### Practical Example: Real-Time Forecasting

For real-time analysis, right alignment is essential. Let's simulate what a
forecaster would have seen at different points in time:

```{r realtime-example}
# Simulate real-time analysis: what would we see in Dec 2022?
cutoff_date <- as.Date("2022-12-31")

# Data available up to cutoff
historical_data <- vehicles |>
  filter(date <= cutoff_date)

# Apply right-aligned MA (what we could compute in real-time)
realtime_ma <- historical_data |>
  augment_trends(
    value_col = "production",
    methods = "ma",
    window = 12,
    align = "right"
  )

# Show last 6 months of trend
realtime_ma |>
  slice_tail(n = 6) |>
  select(date, production, trend_ma)
```

With right alignment, the trend is available immediately as new data arrives,
making it suitable for real-time monitoring dashboards and nowcasting
applications.

### Alignment and Missing Values

Different alignments produce NAs in different locations:

```{r na-pattern}
# Check NA pattern for each alignment
na_summary <- vehicles_align |>
  summarise(
    center_nas = sum(is.na(trend_center)),
    right_nas = sum(is.na(trend_right)),
    left_nas = sum(is.na(trend_left))
  )

na_summary
```

For a 12-month window: - **Center**: \~6 NAs at start and \~6 at end -
**Right**: \~11 NAs at start, none at end (can compute trend up to present) -
**Left**: None at start, \~11 NAs at end

## Exponentially Weighted Moving Average (EWMA)

Unlike simple MA which weights all observations equally, EWMA gives **more
weight to recent observations**. This makes it more responsive to recent
changes.

### How It Works

EWMA uses a smoothing parameter α (alpha) between 0 and 1:

```         
EWMA(t) = α × X(t) + (1 - α) × EWMA(t-1)
```

- Higher α (e.g., 0.7): More responsive to recent data
- Lower α (e.g., 0.1): Smoother, similar to long-window MA

### Comparing MA and EWMA

```{r ewma-comparison}
# Apply both methods separately (EWMA cannot use both window and smoothing)
# First: MA with window parameter
vehicles_ma <- vehicles_recent |>
  augment_trends(
    value_col = "production",
    methods = "ma",
    window = 12
  )

# Second: EWMA with smoothing (alpha) parameter
vehicles_ewma <- vehicles_recent |>
  augment_trends(
    value_col = "production",
    methods = "ewma",
    smoothing = 0.3
  )

# Combine the results
vehicles_ma_ewma <- vehicles_recent |>
  left_join(
    select(vehicles_ma, date, trend_ma),
    by = "date"
  ) |>
  left_join(
    select(vehicles_ewma, date, trend_ewma),
    by = "date"
  )

# Prepare for plotting
plot_data <- vehicles_ma_ewma |>
  select(date, production, trend_ma, trend_ewma) |>
  pivot_longer(
    cols = c(production, trend_ma, trend_ewma),
    names_to = "method",
    values_to = "value"
  ) |>
  mutate(
    method = case_when(
      method == "production" ~ "Original",
      method == "trend_ma" ~ "12-month MA",
      method == "trend_ewma" ~ "EWMA (α=0.3)"
    )
  )

# Plot
ggplot(plot_data, aes(x = date, y = value, color = method)) +
  geom_line(linewidth = 0.9) +
  labs(
    title = "Simple MA vs EWMA",
    subtitle = "EWMA emphasizes recent observations more than simple MA",
    x = "Date",
    y = "Production (thousands of units)",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Choosing Alpha for EWMA

Let's see how different alpha values affect the trend:

```{r ewma-alpha}
# Test different alpha values
alphas <- c(0.1, 0.3, 0.5, 0.8)

vehicles_alphas <- vehicles_recent

for (a in alphas) {
  temp <- vehicles_recent |>
    augment_trends(value_col = "production", methods = "ewma", smoothing = a) |>
    select(trend_ewma)

  names(temp) <- paste0("ewma_", a)
  vehicles_alphas <- bind_cols(vehicles_alphas, temp)
}

# Plot
plot_data <- vehicles_alphas |>
  select(date, production, starts_with("ewma_")) |>
  pivot_longer(
    cols = c(production, starts_with("ewma_")),
    names_to = "method",
    values_to = "value"
  ) |>
  mutate(
    method = case_when(
      method == "production" ~ "Original",
      method == "ewma_0.1" ~ "α = 0.1 (smooth)",
      method == "ewma_0.3" ~ "α = 0.3",
      method == "ewma_0.5" ~ "α = 0.5",
      method == "ewma_0.8" ~ "α = 0.8 (responsive)"
    )
  )

ggplot(plot_data, aes(x = date, y = value, color = method)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "EWMA with Different Alpha Values",
    subtitle = "Higher alpha = more weight on recent data",
    x = "Date",
    y = "Production (thousands of units)",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Guidelines for alpha**: - Smooth trend: α = 0.1 to 0.2 - Balanced: α = 0.3 to
0.4 - Responsive: α = 0.5 to 0.7 - Very responsive: α = 0.8+

## Advanced Moving Averages

The `trendseries` package includes several advanced MA methods designed to
reduce lag while maintaining smoothness.

### Comparing Advanced Methods

```{r advanced-ma}
# Apply multiple advanced MA methods
# Note: EWMA uses smoothing, other methods use window
# Apply window-based methods
vehicles_window_methods <- vehicles_recent |>
  augment_trends(
    value_col = "production",
    methods = c("ma", "wma"),
    window = 12
  )

# Apply EWMA with smoothing parameter
vehicles_ewma_method <- vehicles_recent |>
  augment_trends(
    value_col = "production",
    methods = "ewma",
    smoothing = 0.3
  )

# Combine results
vehicles_advanced <- vehicles_recent |>
  left_join(
    select(vehicles_window_methods, date, starts_with("trend_")),
    by = "date"
  ) |>
  left_join(
    select(vehicles_ewma_method, date, trend_ewma),
    by = "date"
  )

# Prepare for plotting
plot_data <- vehicles_advanced |>
  select(date, production, starts_with("trend_")) |>
  pivot_longer(
    cols = c(production, starts_with("trend_")),
    names_to = "method",
    values_to = "value"
  ) |>
  mutate(
    method = case_when(
      method == "production" ~ "Original",
      method == "trend_ma" ~ "Simple MA",
      method == "trend_ewma" ~ "EWMA",
      method == "trend_wma" ~ "Weighted MA"
    )
  )

# Plot
ggplot(plot_data, aes(x = date, y = value, color = method)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Advanced Moving Average Methods",
    subtitle = "Weighted MA and EWMA reduce lag compared to simple MA",
    x = "Date",
    y = "Production (thousands of units)",
    color = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Method Characteristics

| Method | Smoothness | Responsiveness | Complexity | Best For |
|----|----|----|----|----|
| **MA** | High | Low | Very Simple | Stable trends, teaching |
| **EWMA** | Medium | Medium | Simple | General purpose, recent data matters |
| **Weighted MA** | Medium | Medium | Simple | Emphasizing recent observations |

## Practical Applications

### Application 1: Identifying Trend Changes

Moving averages help identify when trends change direction. Let's look at the
IBC-Br economic activity index:

```{r trend-changes}
# Get recent IBC-Br data
ibcbr_recent <- ibcbr |>
  slice_tail(n = 72)

# Apply EWMA for responsiveness
ibcbr_trend <- ibcbr_recent |>
  augment_trends(
    value_col = "index",
    methods = "ewma",
    smoothing = 0.25
  )

# Prepare plot
plot_data <- ibcbr_trend |>
  select(date, index, trend_ewma) |>
  pivot_longer(
    cols = c(index, trend_ewma),
    names_to = "series",
    values_to = "value"
  ) |>
  mutate(
    series = ifelse(series == "index", "Original", "EWMA Trend")
  )

# Plot
ggplot(plot_data, aes(x = date, y = value, color = series)) +
  geom_line(linewidth = 0.9) +
  labs(
    title = "IBC-Br Economic Activity Index",
    subtitle = "EWMA trend helps identify economic turning points",
    x = "Date",
    y = "Index Value",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Application 2: Seasonal vs Non-Seasonal Data

Moving averages work differently on seasonal data. Let's compare electricity
consumption (seasonal) with vehicle production (less seasonal):

```{r seasonal-comparison}
# Get recent electricity data (seasonal)
electric_recent <- electric |>
  slice_tail(n = 60)

# Apply same 12-month MA to both series
electric_ma <- electric_recent |>
  augment_trends(value_col = "consumption", methods = "ma", window = 12)

vehicles_ma_comp <- vehicles_recent |>
  augment_trends(value_col = "production", methods = "ma", window = 12)

# Create plots
p1 <- electric_ma |>
  select(date, consumption, trend_ma) |>
  pivot_longer(cols = c(consumption, trend_ma), names_to = "series") |>
  mutate(series = ifelse(series == "consumption", "Original", "12-month MA")) |>
  ggplot(aes(x = date, y = value, color = series)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Electricity (Seasonal)",
    x = NULL,
    y = "GWh",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

p2 <- vehicles_ma_comp |>
  select(date, production, trend_ma) |>
  pivot_longer(cols = c(production, trend_ma), names_to = "series") |>
  mutate(series = ifelse(series == "production", "Original", "12-month MA")) |>
  ggplot(aes(x = date, y = value, color = series)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Vehicles (Less Seasonal)",
    x = NULL,
    y = "Thousands",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Display plots
print(p1)
print(p2)
```

**Key insight**: For strongly seasonal data like electricity consumption, a
12-month MA removes the seasonal pattern effectively. For less seasonal data
like vehicle production, the MA primarily smooths out irregular fluctuations.

### Application 3: Cross-Series Comparison

When comparing multiple economic indicators, moving averages help focus on the
underlying trends:

```{r cross-series}
# Prepare data for three indicators
multi_series <- bind_rows(
  ibcbr_recent |>
    select(date, value = index) |>
    mutate(indicator = "Economic Activity"),

  vehicles_recent |>
    select(date, value = production) |>
    mutate(indicator = "Vehicle Production"),

  electric_recent |>
    select(date, value = consumption) |>
    mutate(indicator = "Electricity")
)

# Apply EWMA to all series
multi_trends <- multi_series |>
  group_by(indicator) |>
  augment_trends(
    value_col = "value",
    methods = "ewma",
    frequency = 12,
    smoothing = 0.2
  ) |>
  ungroup()

# Normalize trends to first observation = 100
multi_normalized <- multi_trends |>
  group_by(indicator) |>
  mutate(
    trend_normalized = (trend_ewma / first(trend_ewma)) * 100
  ) |>
  ungroup()

# Plot normalized trends
ggplot(multi_normalized, aes(x = date, y = trend_normalized, color = indicator)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Comparing Economic Indicators: EWMA Trends",
    subtitle = "Normalized to first observation = 100",
    x = "Date",
    y = "Index (normalized)",
    color = "Indicator"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

This reveals how different sectors of the economy moved together or diverged
over time.

## Choosing the Right Moving Average

Here's a practical decision guide:

### Start Here: Basic Questions

1.  **Do you need something simple and interpretable?**
    - → Use **Simple MA** with window = 12 (monthly) or 4 (quarterly)
2.  **Does recent data matter more than old data?**
    - → Use **EWMA** with α = 0.2-0.4
3.  **Is the trend changing quickly and you need to catch it?**
    - → Use **EWMA** with α = 0.5-0.7 or **Zero-Lag EMA**
4.  **Do you need weighted averaging with more emphasis on recent data?**
    - → Use **Weighted MA** or **Zero-Lag EMA**
5.  **Is your data strongly seasonal?**
    - → Consider **STL decomposition** instead (see advanced vignette)

### Parameter Selection Quick Reference

For **monthly data**:

```{r eval=FALSE}
# Conservative (smooth)
data |> augment_trends(value_col = "value", methods = "ma", window = 24)
data |> augment_trends(value_col = "value", methods = "ewma", smoothing = 0.15)

# Balanced (recommended starting point)
data |> augment_trends(value_col = "value", methods = "ma", window = 12)
data |> augment_trends(value_col = "value", methods = "ewma", smoothing = 0.3)

# Responsive (catches changes quickly)
data |> augment_trends(value_col = "value", methods = "ma", window = 6)
data |> augment_trends(value_col = "value", methods = "ewma", smoothing = 0.6)
```

For **quarterly data**:

```{r eval=FALSE}
# Conservative
data |> augment_trends(value_col = "value", methods = "ma", window = 8)

# Balanced
data |> augment_trends(value_col = "value", methods = "ma", window = 4)

# Responsive
data |> augment_trends(value_col = "value", methods = "ewma", smoothing = 0.5)
```

## Common Pitfalls and Solutions

### Pitfall 1: Window Too Small

**Problem**: Trend still looks noisy **Solution**: Increase window size or use
EWMA with lower α

### Pitfall 2: Window Too Large

**Problem**: Trend lags behind recent changes **Solution**: Decrease window
size, use EWMA/DEMA, or try Hull MA

### Pitfall 3: Missing Values at Edges

**Problem**: MA produces NA values at the start/end **Solution**: This is
expected - MAs need complete windows. Use methods like HP filter or Kalman
smoother if you need values at edges.

### Pitfall 4: Using MA on Trending Data

**Problem**: MA doesn't remove overall upward/downward trend **Solution**:
Moving averages extract trends, they don't remove them. If you want to detrend
data, consider first-differencing or HP filter gap analysis.

## Summary

Moving averages are versatile tools for trend extraction:

- **Simple MA**: Best for stable trends and when you need interpretability
- **EWMA**: Great general-purpose choice when recent data is more important
- **Weighted MA/Zero-Lag EMA**: Use when you need both smoothness and
  responsiveness

**Key parameters**: - Window size (for MA): 12 months typical for monthly data -
Alpha (for EWMA): 0.2-0.4 for most applications

**Remember**: Always visualize your results and experiment with parameters. The
"best" method and parameters depend on your specific data and analytical goals.

## Further Reading

- For seasonal data: See the "Advanced Methods" vignette on STL decomposition
- For business cycle analysis: See the "Economic Filters" vignette on HP, BK,
  and CF filters
- For general introduction: See the "Getting Started" vignette

## Appendix: Mathematical Details

For readers interested in the mathematical foundations:

### Simple Moving Average

$$\text{MA}_t = \frac{1}{n} \sum_{i=0}^{n-1} X_{t-i}$$

where $X_t$ is the value at time $t$, and $n$ is the window size.

### Exponentially Weighted Moving Average

$$\text{EWMA}_t = \alpha \cdot X_t + (1-\alpha) \cdot \text{EWMA}_{t-1}$$

where $0 < \alpha \leq 1$ is the smoothing parameter. Alternatively expressed
as:

$$\text{EWMA}_t = \alpha \sum_{i=0}^{\infty} (1-\alpha)^i X_{t-i}$$

This shows EWMA as an infinite weighted sum with exponentially decaying weights.

### Weighted Moving Average

$$\text{WMA}_t = \frac{\sum_{i=0}^{n-1} w_i \cdot X_{t-i}}{\sum_{i=0}^{n-1} w_i}$$

where $w_i$ are the weights (typically $w_i = n-i$, giving more weight to recent
observations) and $n$ is the window size.
